Modular analysis of gene expression data with R

### 2009-11-29 isanote/modular.tex
Added SIB to isanote acknowledgement.
isanote: added Sven's changes.

\title[Modular analysis]{Modular analysis of gene expression data with
  GNU R or Matlab}
\author[G\'abor Cs\'ardi \textit{et~al}]{G\'abor Cs\'ardi\,,$^{1,2}$,
  Zolt\'an Kutalik\,$^{1,2}$ and Sven Bergmann\,$^{1,2}$}
\address{$^{1}$Department of Medical Genetics, and 
  $^{2}$Swiss Institute of Bioinformatics,
  University of Lausanne, Rue de Bugnon 27, CH-1005 Lausanne,
  Switzerland.}

@@

\begin{abstract}

\section{Summary:}
\section{Availability:}
\section{Contact:} \href{Sven.Bergmann@unil.ch}{Sven.Bergmann@unil.ch}
\end{abstract}

@@

% What is biclustering.

Biclustering is the idea of clustering the rows and the columns of a
data matrix, together, usually not independently. In other words, the
goal of a biclustering algorithm is to find blocks in the reordered
data matrix that have, such that the blocks have correlated rows and
columns. It is important to note that these blocks (or biclusters)
have context, i.e. the rows only need to be correlated across the
columns of the block, but not across other columns of the data matrix;
and vice-versa.

When it comes to gene expression data, a bicluster means a subset of
genes ($G$) (or probes, probe sets) and a subset of samples ($S$),
such that genes $G$ are exactly coexpressed across $S$ and samples $S$
have correlated expression profile exactly across $G$.

% What is the ISA.

The Iterative Signature Algorithm (ISA)~\cite{sa,isa,ismod} is a 
powerful biclustering algorithm. It is able to identify biclusters
that overlap and it is resilient to noise.

In this short note, we introduce the \Rpackage{isa2} and
\Rpackage{eisa} GNU R~\cite{R} packages; these implement the ISA
biclustering method, provide a convenient interface to run it using
the BioConductor~\cite{BioC} software package, and also contain
visualization tools.

\section{Methods}

\begin{figure*}
\centering
\includegraphics[width=\textwidth]{isa2workflow3}
\caption{}
\label{fig:workflow}
\end{figure*}

% Batch correction

% Normalization.

% Random and smart seeding.

% The ISA iteration.

% Merging the modules.

% The robustness measure and filtering the modules.

% Module trees.

\section{R Implementation}

Two packages: isa2 does the computation, eisa does everything related
to biology.

\subsection{Work flow}

Two workflows, the simple workflow does everything with default
parameters. In the detailed workflow the user has access to every step
of the analysis.

\subsubsection{The simple work flow}

\subsubsection{The detailed work flow}

\subsection{Visualization}

Expression plots, GO tree plots, module tree plots, HTML summary
generation. ExpressionView.

\subsection{Connection to other software}

The biclust package.

\section{Matlab implementation}

@@

\paragraph{Funding\textcolon}

\paragraph{Conflict of interest\textcolon} None declared.

\bibliographystyle{natbib}
\bibliography{isa}
\nocite{*}

>>>>

\title[Modular analysis]{Modular analysis of gene expression data with R}
\author[G\'abor Cs\'ardi \textit{et~al}]{G\'abor Cs\'ardi\,,$^{1,2}$,
  Zolt\'an Kutalik\,$^{1,2}$ and Sven Bergmann\,$^{1,2}$}
\address{$^{1}$Department of Medical Genetics, and
  $^{2}$Swiss Institute of Bioinformatics,
  University of Lausanne, Rue de Bugnon 27, CH-1005 Lausanne,
  Switzerland.}

@@

\begin{abstract}
\section{Summary:}
Large sets of data, like expression profile from many samples, require
analytic tools to reduce their complexity. Classical (bi-)clustering
algorithms typically attribute elements (genes, arrays) to disjoint groups
(``clusters''). Yet, in some cases overlapping cluster assignments would suit
the biological reality much better.

The Iterative Signature Algorithm (ISA) was designed to overcome this and
other limitations of standard clustering algorithms. It aims to reduce the
complexity of very large sets of data by decomposing it into so-called
``modules''. In the context of gene expression data these modules consist of
subsets of genes that exhibit a coherent expression profile only over a
subset of microarray experiments. Genes and arrays may be attributed to
multiple modules and the level of required coherence can be varied resulting
in different ``resolutions'' of the modular mapping. Since the ISA does not
rely on the computation of correlation matrices (like many other tools), it
is extremely fast even for very large datsets.

In this short note, we introduce two BioConductor~\cite{BioC} software
package written in GNU~R: The~\Rpackage{isa2} includes a highly optimized
implementation of the ISA and~\Rpackage{eisa} provide a convenient interface
to run the ISA and visualize its output.

\section{Availability:}
\href{http://www.unil.ch/cbg/homepage/software.html}%
{http://www.unil.ch/cbg/homepage/software.html}
\section{Contact:} \href{Sven.Bergmann@unil.ch}{Sven.Bergmann@unil.ch}
\end{abstract}

@@

% How the ISA works
The ISA can be applied to identify coherent substructures (i.e. modules)
from any rectangular matrix of data. To be specific we consider here the
case of transcriptomics data corresponding to a selection of gene-expression
profiles from a collection of samples. The method has been described in
detail in Refs~\cite{BioISA,PREISA}. Here only give a brief summary:

The ISA identifies modules by an iterative procedure. The algorithm starts
from an input seed (corresponding to some set of genes or samples) which are
refined at each iteration by adding and/or removing genes and/or samples
until the process converges to a stable set, which is referred to as a
transcription module. This iterative procedure is typically performed to a
large number of seeds. In the unsupervised approach these seeds are chosen
randomly to sample uniformly the immense search space. We also implemented a
semi-supervised method, to which we refer as ``smart seeding'', where the
seeds a biased to start with certain sets of genes or samples for which
there is prior knowledge.

The output of ISA is a collection of potentially overlapping modules. Each
gene and each sample is attributed a score that reflects the strength of the
association with the module.

In the following we first outline how our software tool is commonly for a
modular analysis (Section~\ref{sec:methods}) and then we describe in more
detail the R packages that implement the ISA and visualize its output
(Section~\ref{sec:implementation}).

\section{Methods}%
\label{sec:methods}

\begin{figure*}
\centering
\includegraphics[width=\textwidth]{isa2workflow3}
\caption{Work flow of a typical modular analysis study, with the
  \Rpackage{eisa} package (left); and four bicluster visualization
  techniques (right). The input of the \Rpackagescriptional regulation one often combines several
microarray experiments into a single dataset. In such a case it is important
to properly normalize such data and reduce the bias due to constituent
datasets. Several methods address this challenge,
see e.g.~\cite{johnson07} for an algorithm that has a GNU R implementation.
(S: Do we have a standard flag to call their software?)

% Filtering
\emph{Gene filtering.}
Genes that have very low expression levels in all samples, carry little if
any information and may reflect ineffective array probes, etc. Since these
genes are likely to contribute only noise to the analysis, we integrated the
possibility to remove such genes. (S: Describe how this is done, and if it
requires setting a threshold!)

% Normalization.
\emph{ISA normalization.}
In each iteration of the ISA weighted sums of expression levels over either
genes or samples are computed.  for a number of . Since different genes
typically show different levels of base
expression and variance, it is important to standardize expression
levels to $Z$-scores. The ISA uses two se, the gene
threshold and the sample threshold. ISA modules can be organized into
a directed graph, a module tree, in which there is an edge from module
$A$ to module $B$, if the ISA converges to module $B$ from module $A$,
with the same threshold parameters  that were used to find module
$B$. A module tree provides a hierarchical modular description of a
data set.

\section{Implementation}%
\label{sec:implementation}

The ISA and accompanying visualization tools are implemented in two R
packages. The \Rpackage{isa2} package contains the implementation of
the basic ISA itself; this package can be used to analyze any tabular
data. The \Rpackage{eisa} package builds on \Rpackage{isa2}. It adds
support to standard BioConductor gene expression data structures; and
contains gene expression specific visualization tools, see some of
these in Fig.~\ref{fig:workflow}.

Both the \Rpackage{isa2} and \Rpackage{eisa} packages support two
workflows. The \emph{simple workflow} involves a single R function
call and runs all ISA steps (1-3 on Fig.~\ref{fig:workflow}.) with
their default parameters.

The \emph{detailed workflow} means running every step of the modular
analysis separately, possibly with non-default parameters. This allows
the users to tailor the ISA completely according their needs.

For users that prefer Matlab over R, we created a Matlab package for
ISA, please see the ISA homepage for more information.

% \subsubsection{Visualization}

The \Rpackage{eisa} package implements a set of visualization
techniques for biclusters, some of these are specific to ISA, others
are not. Please see Fig.~\ref{fig:workflow} for some of these.

The \Rpackage{ExpressionView} R package implements an interactive tool
for visualizing ISA (or other) overlapping biclusters.

% \subsubsection{Connection to other software}

The \Rpackage{biclust} package, \cite{biclust}, implements a number of
biclustering algorithms in a unified framework. The \Rpackage{eisa}
package include tools to convert between \Rpackage{biclust} and ISA
biclusters. This allows the cross-talk of the functions in the two
packages.

@@

\paragraph{Funding\textcolon} The authors are greatful to the Swiss
Institute of Bioinformatics for support.
\paragraph{Conflict of interest\textcolon} None declared.

\bibliographystyle{natbib}
\bibliography{isa}
